---
title: "Gender Pay Gap in UK Universities"
author: "Group 5: Benjamin Bray, Leanne Allen, Nadia Korchynska and Stephen Wotton"
date: "2021-10-27"
output:
  html_document:
    theme: yeti
    number_sections: false
fontsize: 11pt
urlcolor: blue
header-includes:
  - \usepackage{dplyr}
  - \usepackage{tidyverse}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE,
                      fig.align="center"
                    )
```
```{r generate_dataset_for_analysis}
library(dplyr)
library(tidyverse)

# cater for running as script and as markdown doc using knit - directory differs in intellij
data_path <-  "../data"
if(!dir.exists(data_path)){
  data_path <- "data"
}

# define function for reading table and adding the start and end data to the data
read_data_augment_with_years <- function(file_to_read){
  # find the years from file name
  years_result <- gregexpr("[0-9]{4}", file_to_read)
  start_year_index <- years_result[[1]][[1]]
  end_year_index <- years_result[[1]][[2]]
  start_year <- substr(file_to_read, start_year_index, start_year_index + 3)
  end_year <- substr(file_to_read, end_year_index, end_year_index + 3)

  # read table and add the years to data
  table <- read.csv(file_to_read) %>%
    mutate(start_year = start_year) %>%
    mutate(end_year = end_year)

  table
}

#read all the data and merge together
data_files <- list.files(path=data_path, full.names=TRUE, pattern="^UK Gender Pay Gap Data -(.*).csv$" )
data_tables <- lapply(data_files,  read_data_augment_with_years)
merged_tables <- Reduce(function(x,y) {merge(x,y, all=TRUE)}, data_tables)

# Join the university list with the pay gap data
# Note that we can't join on name as they are not consistent
# we are told so must change to join on ID
uni <- read.csv(str_interp("${data_path}/universities.csv"))
uni_data_find_employers <- inner_join(uni, merged_tables, by="EmployerName")
uni_employers_by_id <-  uni_data_find_employers %>%
                        select(EmployerId, EmployerName) %>%
                        distinct()

# We join on name, find the id from the name then join by id which will
# return only the employers of interest in a consitent manner i.e. universities
# Exclude current year as only data for 2 universities, so not representative sample.
# Resolve names from join renaming the column from the universities file and
# dropping the other with select(-starts_with....) which returns all but that column
uni_data_emp <- inner_join(uni, uni_employers_by_id, by="EmployerName")
uni_data <- inner_join(uni_data_emp, merged_tables, by="EmployerId") %>%
            filter( start_year !=2021) %>%
            rename("EmployerName"=EmployerName.x) %>%
            select(-starts_with("EmployerName.y"))

# Find the universities which have full result sets for the years of interest i.e. 4 years pre start year 2021
uni_with_full_result_sets <- aggregate(uni_data$EmployerId,
                                        by=list(uni_data$EmployerId, uni_data$institution), FUN=length) %>%
                            select(EmployerId=Group.1,institution=Group.2, Count=x  ) %>%
                            filter(Count == 4) %>%
                            inner_join(uni_data, by="EmployerId")


```
## Plots
```{r output_uni_names, include = TRUE}
  uni_with_full_result_sets %>%
          select(EmployerName) %>%
          distinct() %>%
          arrange(EmployerName)
```

## Data Sources
Data was obtained from two sources:

### Sheffield University
The department provided a set of University names split by whether they were a university prior to 1992 or were former polytechnics which gained university status post 1992.

### UK Gender Pay Gap Government Service
The UK Government provides pay gap data by gender from companies which are obligated and those which elect to submit it.

The data service is at: https://gender-pay-gap.service.gov.uk/

Note that the data diction ary can be found on page https://gender-pay-gap.service.gov.uk/viewing/download by clicking on the "About CSV data" link (which is a HTML details element without an ID hence the report is unable to link to it for you).

### Further Notes
It should be noted that twenty-seven Universities changed their employer name  throughout the data which required data to be cleaned.

In data submitted for 2021/22 when the data was downloaded it contained only two universities; thus this year was excluded as the dataset was incomplete.

Of the ninety-six universities within the datasets only eighty-nine had data spanning all fours years of interest, namely: 2017/18, 2018/19, 2019/20, 2020/21.
