---
title: "Case Study 1 - University Gender Pay Gap"
author: "Leanne Allen, Benjamin Bray, Nadiia Korchynska, Stephen Wotton"
date: "1/11/2021"
output:
html_document:
  theme: yeti
word_document: default
pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE)

```

```{r load_data, include = FALSE, echo=FALSE}
#load data
library(tidyverse)

pay17 <- read_csv("data/UK Gender Pay Gap Data - 2017 to 2018.csv")
pay18 <- read_csv("data/UK Gender Pay Gap Data - 2018 to 2019.csv")
pay19 <- read_csv("data/UK Gender Pay Gap Data - 2019 to 2020.csv")
pay20 <- read_csv("data/UK Gender Pay Gap Data - 2020 to 2021.csv")
universities <- read_csv("data/universities.csv")

pay17year <- pay17 %>%
  mutate(year = "2017/18")

pay18year <- pay18 %>%
  mutate(year = "2018/19")

pay19year <- pay19 %>%
  mutate(year = "2019/20")

pay20year <- pay20 %>%
  mutate(year = "2020/21")

mergedPayGapData <- rbind(pay17year, pay18year, pay19year, pay20year)

# Join the university list with the pay gap data
# Note that we can't join on name as they are not consistent
# we are told so must change to join on ID
universitiesWithEmployerNameInitialMatch <- inner_join(universities, mergedPayGapData, by="EmployerName")
universityEmployersById <-  universitiesWithEmployerNameInitialMatch %>%
  select(EmployerId, EmployerName) %>%
  distinct()

# We join on name, find the id from the name then join by id which will
# return only the employers of interest in a consitent manner i.e. universities
# Exclude current year as only data for 2 universities, so not representative sample.
# Resolve names from join renaming the column from the universities file and
# dropping the other with select(-starts_with....) which returns all but that column
payGapDataAllUniversities <- inner_join(universities, universityEmployersById, by="EmployerName") %>%
  inner_join(mergedPayGapData, by="EmployerId") %>%
  rename("EmployerName"=EmployerName.x) %>%
  select(-starts_with("EmployerName.y"))

# Find the universities which have full result sets for the years of interest i.e. 4 years data
# Note there are 89 universities with full set of data for 4 years, therefore we need 89*4 = 356 rows
# previously we were only showing 321 rows
payGapData <- aggregate(payGapDataAllUniversities$EmployerId,
                                       by=list(payGapDataAllUniversities$EmployerId, payGapDataAllUniversities$institution), FUN=length) %>%
  select(EmployerId=Group.1,institution=Group.2, Count=x  ) %>%
  filter(Count == 4) %>%
  inner_join(payGapDataAllUniversities, by="EmployerId") %>%
  rename("institution"=institution.x) %>%
  select(-starts_with("institution.y"))


```


```{r analysis_diff_median, include=FALSE, echo = FALSE}
payGapData %>%
  select(EmployerName, year, DiffMedianHourlyPercent) %>%
  pivot_wider(names_from = year,
              values_from = DiffMedianHourlyPercent) %>%
  na.omit()
```

# Table #1

```{r analysis_paygap, echo = FALSE}

hourly<-payGapData %>%
  group_by(year, institution) %>%
  summarise(meanGap = mean(DiffMedianHourlyPercent), .groups = "drop") %>%
  pivot_wider(names_from = institution, values_from = meanGap)
knitr::kable(hourly, digits = 2, align = "cc", col.names = c("Year", "Post-92", "Pre-92"))

payGapDataLong <- payGapData %>%
  pivot_longer(DiffMeanHourlyPercent:FemaleTopQuartile,
               names_to = "payGapMeasure",
               values_to = "value") %>%
  select(EmployerName, year, payGapMeasure, value, institution)

```



# Table #2 - Average Median Gender Percentage Difference For Hourly Wage

```{r analysis_percent_difference, echo = FALSE}
hourly<-payGapData %>%
  group_by(year) %>%
  summarise(meanGap = mean(DiffMedianHourlyPercent))
knitr::kable(hourly, col.names = c("Year", "Percentage Difference"), digits = 2, align = "cc")
```


# Plot #1 - Pre-92 and Post-92 Universities Average Gender Difference Bonus Percentages


```{r analysis_bonus, fig.align = 'center', fig.cap = "PlotAcross all four years of data, it is consistent that Pre-92 universities have a higher percentage gap in the mean percentages of bonuses close to 50% or above, whereas Post-92 universities remain with a gap < 25%. The source of the data is from... ", out.width = "75%", , message=FALSE, warning = FALSE}


payGapData%>%
  ggplot(aes(y= DiffMeanBonusPercent, x = institution))+geom_boxplot()+facet_grid(cols = vars(year)) + labs(y = "Average Gender Difference of Bonus Percentages", x = "") + theme_minimal()+theme(strip.background = element_rect(fill="#FC4E07" ))+ scale_x_discrete(breaks = c("post-92", "pre-92"), labels =c( "Post-92", "Pre-92"))

```

## Data Sources
Data was obtained from two sources:

### Sheffield University
The department provided a set of University names split by whether they were a university prior to 1992 or were former polytechnics which gained university status post 1992.

### UK Gender Pay Gap Government Service
The UK Government provides pay gap data by gender from companies which are obligated and those which elect to submit it.

The data service is at: https://gender-pay-gap.service.gov.uk/

Note that the data diction ary can be found on page https://gender-pay-gap.service.gov.uk/viewing/download by clicking on the "About CSV data" link (which is a HTML details element without an ID hence the report is unable to link to it for you).

### Further Notes
It should be noted that twenty-seven Universities changed their employer name  throughout the data which required data to be cleaned.

In data submitted for 2021/22 when the data was downloaded it contained only two universities; thus this year was excluded as the dataset was incomplete.

Of the ninety-six universities within the datasets only eighty-nine had data spanning all fours years of interest, namely: 2017/18, 2018/19, 2019/20, 2020/21.




