---
title: "Case Study 1 - University Gender Pay Gap"
author: "Leanne Allen, Benjamin Bray, Nadiia Korchynska, Stephen Wotton"
date: "1/11/2021"
output:
html_document:
  theme: yeti
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE)

```

```{r load_data, include = FALSE, echo=FALSE}
#load data
library(tidyverse)

pay17 <- read_csv("../data/UK Gender Pay Gap Data - 2017 to 2018.csv")
pay18 <- read_csv("../data/UK Gender Pay Gap Data - 2018 to 2019.csv")
pay19 <- read_csv("../data/UK Gender Pay Gap Data - 2019 to 2020.csv")
pay20 <- read_csv("../data/UK Gender Pay Gap Data - 2020 to 2021.csv")
universities <- read_csv("../data/universities.csv")

pay17year <- pay17 %>%
  mutate(year = "2017/18")

pay18year <- pay18 %>%
  mutate(year = "2018/19")

pay19year <- pay19 %>%
  mutate(year = "2019/20")

pay20year <- pay20 %>%
  mutate(year = "2020/21")

mergedPayGapData <- rbind(pay17year, pay18year, pay19year, pay20year)

# Join the university list with the pay gap data
# Note that we can't join on name as they are not consistent
# we are told so must change to join on ID
universitiesWithEmployerNameInitialMatch <- inner_join(universities, mergedPayGapData, by="EmployerName")
universityEmployersById <-  universitiesWithEmployerNameInitialMatch %>%
  select(EmployerId, EmployerName) %>%
  distinct()

# We join on name, find the id from the name then join by id which will
# return only the employers of interest in a consitent manner i.e. universities
# Exclude current year as only data for 2 universities, so not representative sample.
# Resolve names from join renaming the column from the universities file and
# dropping the other with select(-starts_with....) which returns all but that column
payGapDataAllUniversities <- inner_join(universities, universityEmployersById, by="EmployerName") %>%
  inner_join(mergedPayGapData, by="EmployerId") %>%
  rename("EmployerName"=EmployerName.x) %>%
  select(-starts_with("EmployerName.y"))

# Find the universities which have full result sets for the years of interest i.e. 4 years data
# Note there are 89 universities with full set of data for 4 years, therefore we need 89*4 = 356 rows
# previously we were only showing 321 rows
payGapData <- aggregate(payGapDataAllUniversities$EmployerId,
                        by=list(payGapDataAllUniversities$EmployerId, payGapDataAllUniversities$institution), FUN=length) %>%
  select(EmployerId=Group.1,institution=Group.2, Count=x  ) %>%
  filter(Count == 4) %>%
  inner_join(payGapDataAllUniversities, by="EmployerId") %>%
  rename("institution"=institution.x) %>%
  select(-starts_with("institution.y"))


```


```{r analysis_diff_median, include=FALSE, echo = FALSE}
payGapData %>%
  select(EmployerName, year, DiffMedianHourlyPercent) %>%
  pivot_wider(names_from = year,
              values_from = DiffMedianHourlyPercent) %>%
  na.omit()
```

## Table #1

```{r analysis_paygap, echo = FALSE}

hourly<-payGapData %>%
  group_by(year, institution) %>%
  summarise(meanGap = mean(DiffMedianHourlyPercent), .groups = "drop") %>%
  pivot_wider(names_from = institution, values_from = meanGap)
knitr::kable(hourly, digits = 2, align = "cc", col.names = c("Year", "Post-92", "Pre-92"))

payGapDataLong <- payGapData %>%
  pivot_longer(DiffMeanHourlyPercent:FemaleTopQuartile,
               names_to = "payGapMeasure",
               values_to = "value") %>%
  select(EmployerName, year, payGapMeasure, value, institution)

```



## Table #2 - Average Median Gender Percentage Difference For Hourly Wage

```{r analysis_percent_difference, echo = FALSE, message=FALSE}
hourly<-payGapData %>%
  group_by(year) %>%
  summarise(meanGap = mean(DiffMedianHourlyPercent))
knitr::kable(hourly, col.names = c("Year", "Percentage Difference"), digits = 2, align = "cc")
```


## Plot #1 - Pre-92 and Post-92 Universities Average Gender Difference Bonus Percentages


```{r analysis_bonus, fig.align = 'center', fig.cap = "PlotAcross all four years of data, it is consistent that Pre-92 universities have a higher percentage gap in the mean percentages of bonuses close to 50% or above, whereas Post-92 universities remain with a gap < 25%. The source of the data is from... ", out.width = "75%", , message=FALSE, warning = FALSE}


payGapData%>%
  filter(is.na(DiffMeanBonusPercent) == FALSE) %>%
  ggplot(aes(y= DiffMeanBonusPercent, x = institution))+geom_boxplot()+facet_grid(cols = vars(year)) + labs(y = "Average Gender Difference of Bonus Percentages", x = "") + theme_minimal()+theme(strip.background = element_rect(fill="#FC4E07" ))+ scale_x_discrete(breaks = c("post-92", "pre-92"), labels =c( "Post-92", "Pre-92"))

```
### Data inconsistencies

Note that there were thirty-six rows of data which did not have the column `DiffMeanBonusPercent` defined hence we have excluded those results.  These were from seventeen different universities:

|Number of Universities Missing Data Values|Number of Rows of Missing Values|
|:----------------:|:----------------:|
|5|1|
|6|2|
|5|3|
|1|4|
Total Universities: 17 | Total Rows: 36|


## Plot #2 - Hourly Pay Overview

```{r analysis_hourly_pay_quartiles, fig.align = 'center', fig.cap = "Although the trend is gradually decreasing, males dominate in the top hourly pay. Females dominate in the lower 3 quartiles of hourly pay with a trend: as pay increases the % of females in the corresponding quartile decreases. This trend is apparent across both categories of institutions, although it is more pronounced in the pre-92 group where males comprised 58.6% of the top quartile in the pre-92 group in 2020/21 compared to 51.1% in the post-92 group. This might be partially explained by there being a greater proportion of males in the pre-92 group in general compared to the post-92 group (45.6% are males in the pre-92 group vs 42.1% in the post-92 group).", out.width = "75%", fig.height = 6 , message=FALSE, warning = FALSE}

## reshaping data. Grouping by year and institution type, taking mean from all the quartiles and rounding it up.
df2 <- select(payGapData, institution, year, MaleLowerQuartile ,FemaleLowerQuartile,
              MaleLowerMiddleQuartile,  FemaleLowerMiddleQuartile, MaleUpperMiddleQuartile,
              FemaleUpperMiddleQuartile, MaleTopQuartile,FemaleTopQuartile) %>%
  group_by(institution , year) %>%
  summarise(mean_MaleLowerQuartile = (round( mean(MaleLowerQuartile), digits = 1)) ,
            mean_FemaleLowerQuartile = round(mean(FemaleLowerQuartile), digits = 1),
            mean_MaleLowerMiddleQuartile = round(mean(MaleLowerMiddleQuartile), digits = 1),
            mean_FemaleLowerMiddleQuartile = round(mean(FemaleLowerMiddleQuartile), digits = 1),
            mean_MaleUpperMiddleQuartile = round(mean(MaleUpperMiddleQuartile), digits = 1),
            mean_FemaleUpperMiddleQuartile = round(mean(FemaleUpperMiddleQuartile), digits = 1),
            mean_MaleTopQuartile = round(mean(MaleTopQuartile), digits = 1),
            mean_FemaleTopQuartile = round(mean(FemaleTopQuartile), digits = 1))

# pivoting data so that all the values appear in one column
df_for_plot <- pivot_longer(df2,
                            cols = -c('institution','year'),
                            names_to = 'gender_quartile',
                            values_to = 'mean') %>%
  mutate(gender = case_when(grepl("Male", gender_quartile) ~ "Male",
                            grepl("Female", gender_quartile) ~"Female")) %>%
  mutate(quartile = case_when(grepl("Male", gender_quartile) ~ gsub("mean_Male","",gender_quartile),
                              grepl("Female", gender_quartile) ~ gsub("mean_Female","",gender_quartile)))

#turning quartiles to factors to force their order in the graph
df_for_plot$quartile <- factor(df_for_plot$quartile, levels=c('TopQuartile','UpperMiddleQuartile','LowerMiddleQuartile','LowerQuartile'))

p2 <- ggplot(df_for_plot, aes(x = year, y = mean, fill = gender)) +
  geom_col(position = 'stack') +
  facet_grid( quartile ~ institution) +
  geom_text(aes(label = mean), size = 3, hjust = 0.5, vjust = 3, position =     "stack") +
  labs(x = "Year", y= "Hourly pay per gender %")
p2

#Plot 3 - Grouping by Post and Pre 92 institutions and assessing their records over the time period

p92 <- payGapDataAllUniversities %>% group_by(institution, year) %>%
  summarise(mean_pct_diff = mean(DiffMedianHourlyPercent))

ggplot(data = p92,
       aes(x = year, y = mean_pct_diff, group = institution,
           colour = institution)) +
  geom_point() +
  labs(x = "Year",
       y = "Mean % Difference (Median Hourly Pay)",
       title = "Mean of percent differences in median hourly pay",
       subtitle = 'By type of University Institution') +
  geom_line() +
  ylim(0,20)


```

## Data Sources
Data was obtained from two sources:

### Sheffield University
The department provided a set of University names split by whether they were a university prior to 1992 or were former polytechnics which gained university status post 1992.

### UK Gender Pay Gap Government Service
The UK Government provides pay gap data by gender from companies which are obligated and those which elect to submit it.

The data service is at: https://gender-pay-gap.service.gov.uk/ The data was downloaded on 30 October 2021 at 12:04pm British Summer Time.

Note that the data diction ary can be found on page https://gender-pay-gap.service.gov.uk/viewing/download by clicking on the "About CSV data" link (the website design stops us from linking to it).

### Further Notes
It should be noted that twenty-seven Universities changed their employer name  throughout the data which required data to be cleaned.

In data submitted for 2021/22 when the data was downloaded it contained only two universities; thus this year was excluded as the dataset was incomplete.

Of the ninety-six universities within the datasets only ninety universities had data spanning all fours years of interest, namely: 2017/18, 2018/19, 2019/20, 2020/21.