---
title: "MAS61002 Project Part B  \n Registration Number: 210138744"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(dplyr)
library(ggplot2)
data <- load("Data/E_PartB_data.RData")
```

# Question 1

# Question 1.1
```{r include=TRUE, results='asis', figures-side, fig.show="hold", out.width="50%"}
library(knitr) # for prettier tables
library(pander) #for adding pandoc headers
library(ggfortify) # so we can use fortify to create a datatable required for knitr
options(digits = 3) # let's keep output to sensible number of places
trialA <- q1dat %>% filter(drug == "A")
trialB <- q1dat %>% filter(drug == "B")
trialA.sv <- Surv(time=trialA$time, event=trialA$status, type="right")
trialASurv <- survfit(trialA.sv ~ 1, data = trialA)
tableA <- fortify(trialASurv) %>% rename(survival=surv, "lower 95% CI"=lower, "upper 95% CI"= upper) %>%
            select(1, 2, 3, 5, 6, 8, 7)
pandoc.header(list("Kaplan-Meier Estimate of Survivor Function for trial drug A"), level = 2)
print(kable(tableA))
trialB.sv <- Surv(time=trialB$time, event=trialB$status, type="right")
trialBSurv <- survfit(trialB.sv ~ 1, data = trialB)
tableB <- fortify(trialBSurv) %>% rename(survival=surv, "lower 95% CI"=lower, "upper 95% CI"= upper)%>%
            select(1, 2, 3, 5, 6, 8, 7)
pandoc.header(list("Kaplan-Meier Estimate of Survivor Function for trial drug B"), level = 2)
print(kable(tableB))

par(mar = c(4, 4, .1, .1))
plot(trialASurv, xlab ="months", ylab = "probablity of survival")
title("Kaplan-Meier  plot of trial A drug", line= -1 )
plot(trialBSurv, xlab ="months", ylab = "probablity of survival")
title( "Kaplan-Meier  plot of trial B drug", line= -1 )


```


# Question 1.2
Things to comments on: earliest death, average death, calculate rough quartiles. longest death. Look at error bounds. Number of people in the study. number who were censored.

# Question 1.3
Answer to Question 1.3 here

# Question 2

# Question 2.1
```{r include=TRUE}
lungcancer.sv <- Surv(time=q2dat$time, event=q2dat$stat, type="right")
model <- lungcancer.sv ~ as.factor(trea) + as.factor(sex) + as.factor(exer) + as.factor(smok) + age
lungcancer.regexp <- survreg(model, dist="exponential", data=q2dat)
summary(lungcancer.regexp)
```
Therefore the accelerated failure time model with an exponential failure time distribution and explanatory variables **x** is:

$$ \begin{aligned}
S(t;\mathbf{x}) &= exp\{{-t\: exp(-(\beta_0 + \beta_1x_{1} + \beta_2x_{2} + \beta_3x_{3} + \beta_4x_{4} + \beta_5x_{5}))}\}\\
&= exp\{{-t\: exp(-(\beta_0 + \beta_1x_{trea} + \beta_2x_{sex} + \beta_3x_{exer} + \beta_4x_{smok} + \beta_5x_{age}))}\}\\
&= exp\{{-t\: exp(-(0.5017 + 0.8144x_{trea} - 0.5328x_{sex} - 0.2566x_{exer} - 0.8821x_{smok} - 0.0113x_{age}))}\}\\
&= exp\{{-t\: exp(-0.5017 - 0.8144x_{trea} + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} + 0.0113x_{age})}\}\\
\end{aligned}
$$

Variable     Coding
--------    -------------------------------------------------------------
sex         0 = male; 1 = female
exer        0 = some exercise; 1 = rarely exercise
smok        0 = not current smoker; 1 = current smoker
trea        0 = treatment A; 1 = treatment B
age         age of patient centred on 50 years (i.e. 60 years old is 10)
t           survival time (years)

# Question 2.2
Given the coding then the model will include the treatment variable for those on treatment B and exclude it for those on treatment A. Let $S_a(t;\mathbf{x})$ be the model for treatment A and $S_b(t;\mathbf{x})$ the model for treatment B, then this gives:

$$ \begin{aligned}
S_a(t;\mathbf{x}) &= exp\{{-t\: exp(-0.5017 + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} + 0.0113x_{age})}\}\\
S_b(t;\mathbf{x}) &= exp\{{-t\: exp(-0.5017 - 0.8144x_{trea} + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} + 0.0113x_{age})}\}\\
\end{aligned}
$$

As $z = \beta_4 /(std. error) = 4.29$ and $4.29 > 1.96$ then there is overwhelming evidence that the treatment has an effect. Considering the proprtional hazards related to treatment A and B we know that $h(t; x |x_1=1)/h(t; x | x_1=0) = exp\:\beta_4$.  As $\beta_4 > 0$ then treatment B ($x_1 = 1$) increases the hazard relative to treatment A ($x_1 = 0$).

$$\begin{aligned}
\text{The 95}\% \text{ confidence interval for treatment }\beta_4\text{ is: }&= \hat{\beta_4} \pm 1.96 \times s.e(\hat{\beta_4}) \\
&= 0.8144 \pm 1.96 \times 0.1898 \\
&= (0.44, 1.19)
\end{aligned}
$$

The hazard function $h(t; \mathbf{x}) = exp\{\mathbf{B}'x\}h_0(t)$ where $h_0(t) is the baseline hazard function.  Let $h_a(t; \mathbf{x})$ and $h_b(t; \mathbf{x})$ be the respective hazard functions of treatment A and treatment B, which are:
$$
\begin{aligned}
h_a(t;\mathbf{x}) &= h_0(t)exp\{\beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_4x_4 + \beta_5x_5 \}\\
h_b(t;\mathbf{x}) &= h_0(t)exp\{{-t\: exp(-0.5017 + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} - 0.8144x_{trea} + 0.0113x_{age})}\}\\
\end{aligned}
$$

Then the hazard ratio of treatment A versus treatment B is given by:
$$
\begin{aligned}
\frac{h_b(t;\mathbf{x})}{h_a(t;\mathbf{x})}  &= \frac{h_0(t)exp\{{-t\: exp(-0.5017 + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} - 0.8144x_{trea} + 0.0113x_{age})}\}}{h_0(t)exp\{{-t\: exp(-0.5017 + 0.5328x_{sex} + 0.2566x_{exer} + 0.8821x_{smok} + 0.0113x_{age})}\}}\\
 &= \frac{h_o(t)exp\{\beta_1\times 1\}}{h_0(t)exp\{\beta_1\times 0\}} \\
 &=exp\{\beta_1\}\\
 &= exp\{0.8144\} \\
 &= 2.26 \\
\end{aligned}
$$
The 95$\%$ confidence interval for the hazard ratio of treatment A versus treatment B ($exp\{\beta_1\}$) is given by:
$$
\begin{aligned}
95\% \text{ CI } &= (exp\{\beta_1 - 1.96*s.e(\beta_1)\}, exp\{\beta_1 + 1.96*s.e(\beta_1)\})\\
&= (exp\{0.8144 - 1.96*0.1898\}, exp\{0.8144 + 1.96*0.1898\})\\
&= (1.56, 3.28)
\end{aligned}
$$
Interpreting the confidence interval of (1.56, 3.28) means that at the $95\%$ confidence level that the hazard for patients on treatment B could be and where between just over one and half times and just over three and a quarter times that of the hazard for treatment A.

# Question 2.3
Answer to Question 2.3 here

# Question 2.4
Answer to Question 2.4 here


